---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Set the working directory for the project.  
```{r echo=T}
setwd("~/datasciencecoursera/RepData_PeerAssessment1")
```
Load the Activity data:
```{r echo=T}
data<-read.csv("activity.csv", header=T, stringsAsFactors = F)
str(data)
```
Transform the date variable from character to a proper Date type:
```{r echo=T}
data$date<-as.Date(data$date, format="%Y-%m-%d")
str(data)
```
Search for missing values in the numeric column and filter out rows with missing data:
```{r echo=T}
table(is.na(data$steps))
#Filter out the missing data:
data_NoNa<-data[complete.cases(data),]
nCompleteCases<-nrow(data_NoNa)
```
The number of complete rows in the data subset that does not contain missing values is:
```{r echo=T}
nCompleteCases
```

Preview of data rows in the clean data set:
```{r echo=T}
head(data_NoNa, 15)
```
## What is mean total number of steps taken per day?
### Total number of steps taken per day:
```{r echo=T, warning=F, message=F}
library(dplyr)
TotalStepsByDate<- data_NoNa %>% group_by(date) %>% summarize(totalSteps = sum(steps))
```
Histogram of Total Number of Steps taken:
```{r echo=T}
hist(TotalStepsByDate$totalSteps, breaks=30, xlab="Bins", main="Histogram: Total Steps by Date")
```

### Mean of the Total Number of Steps take by day:
```{r echo=T}
mean(TotalStepsByDate$totalSteps)
```  

### Median of the Total Number of Steps take by day:
```{r echo=T}
median(TotalStepsByDate$totalSteps)
```
## What is the average daily activity pattern?
Average the steps taken by interval across all days:
```{r, echo=T}
AvgStepsByInterval<- data_NoNa %>% group_by(interval) %>% summarize(avgSteps=mean(steps))
plot(AvgStepsByInterval$interval, AvgStepsByInterval$avgSteps, type="l", xlab="5-Min Interval", ylab="Avg Steps", main="Interval Activity averaged across all days")
```

The 5-minute interval, across all days, with the maximum number of steps is: interval **835**
```{r, echo=T}
AvgStepsByInterval[which.max(AvgStepsByInterval$avgSteps),]
```

## Imputing missing values
The original Activity data set had **`r table(is.na(data$steps))[2]`** missing values in the step variable.
We will fill out the missing values by calculating and assigning the 5-minute interval average number of steps taken across all days, based on the 5-minute interval associated with the missing value.
```{r, echo=T}
Lookup_AvgStepsByInterval<- data_NoNa %>% group_by(interval) %>% summarize(avgSteps=mean(steps))
imputedValuesDataset <- data 
for (i in 1:nrow(imputedValuesDataset)) {
  if (is.na(imputedValuesDataset$steps[i])) {
    imputedValuesDataset$steps[i] <- Lookup_AvgStepsByInterval[which(imputedValuesDataset$interval[i] == Lookup_AvgStepsByInterval$interval), ]$avgSteps
  }
}
```

A preview of the new data set with imputed values is shown below:
```{r, echo=T}
head(imputedValuesDataset, 10)
```

Summarize the new Total Number of Steps taken, grouped by day:
```{r, echo=T}
newTotalStepsByDate<- imputedValuesDataset %>% group_by(date) %>% summarize(totalSteps = sum(steps))
head(newTotalStepsByDate, 10)
```

Create the histogram of the new Total Number of steps taken (data set with imputed values)
```{r, echo=T}
hist(newTotalStepsByDate$totalSteps, breaks=30, xlab="Bins", main="Histogram: Total Steps by Date (Imputed Data Set)")
```

The new mean of the filled data set is:
```{r, echo=T}
mean(newTotalStepsByDate$totalSteps)
```

The mean of the filled data set without missing values was:
```{r, echo=T}
mean(TotalStepsByDate$totalSteps)
```

The means are the same.
     
The new median of the filled data set is:
```{r, echo=T}
median(newTotalStepsByDate$totalSteps)
```

The median of the data set without missing values was:
```{r, echo=T}
median(TotalStepsByDate$totalSteps)
```

The median of the data set with imputed values is slightly bigger.

## Are there differences in activity patterns between weekdays and weekends?
Define a function that will calculate is a date is a weekday or a weekend day.

```{r, echo=T}
determineWeekdayFromWeekend<-function(aDate){
  if( tolower(weekdays(aDate)) %in% c("monday", "tuesday", "wednesday","thursday","friday")) return("weekday")
  if( tolower(weekdays(aDate)) %in% c("saturday", "sunday")) return("weekend")
}
```

Create a new Factor varible in the filled data set that classifies a Date as weekday or weekend:
```{r, echo=T}
imputedValuesDataset$dayType<-as.factor(sapply(imputedValuesDataset$date, determineWeekdayFromWeekend))
```

Calculate the average number of steps taken by 5-minute interval, grouped by weekday or weekend:
```{r, echo=T, message=F}
fiveMinAvgStepsByWeekendOrWeekday<-imputedValuesDataset%>%group_by(dayType,interval)%>% summarize(avgSteps=mean(steps))
str(fiveMinAvgStepsByWeekendOrWeekday)
```

Plot the aggregation
```{r, echo=T, message=F, warning=F}
library(ggplot2)

ggplot(fiveMinAvgStepsByWeekendOrWeekday, aes(interval, avgSteps))+
    geom_line()+
    facet_grid(dayType~.)+
    ggtitle("5-Minute Interval Average Number of Steps, by Day type")+
    ylab("Average 5-min Interval Steps")
```


>Prepared by etsibert@hotmail.com, 03/06/2016



